Optimal Finite Element Integration

### 2015-08-03 TOMS_Paper/assembly.tex
Rewrite perf eval sect intro

Experiments were run on a single core of an Intel architecture, a Sandy Bridge I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache. The Intel \texttt{icc 14.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost, -xAVX, -ip}.

We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods. In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms. The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{ufl-code}.

We evaluate the \textit{speed ups} achieved by three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied. In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning as described in Section~\ref{sec:autotune}). The \texttt{auto} values do not include the autotuner cost, which is commented aside in Section~\ref{sec:auto-analysis}.

%This way, we also highlight the advances achieved over our previous work. 
%...TODO...: dire che i tempi sono la media di 3 runs

The values that we report include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the ``noise'' due to any operations that are not part of the element matrix evaluation itself. However, it has been reiterated over and over (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant. All codes were executed in the context of the Firedrake framework.

We vary several aspects of each form, which follows the approach and the notation of~\cite{quadrature1} and~\cite{Francis}
\begin{itemize}
\item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item The polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
On the other hand, other aspects are fixed 
\begin{itemize}
\item The space of both basis and coefficient functions is Lagrange
\item The mesh is three-dimensional, made of tetrahedrons, for a total of 4374 cells
\end{itemize}

Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will be deeply commented in the next section, must be read as ``plots, or grids, of plots''. Each grid (figure) has two logical axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot in a grid shows speed ups for $[q=1, p=1]$; the plot on its right does the same for $[q=1, p=2]$, and so on. The diagonal of the grid shows plots for which basis and coefficient functions have same polynomial order, that is $q=p$. Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.

A plot reports speed-ups over non-optimized FEniCS-Form-Compiler-generated code. There are three groups of bars, each group referring to a particular version of the code (\texttt{ffc, fix, auto}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

@@

\paragraph{Poisson}
%\begin{figure}[t]  
%...
%%\includegraphics[scale=0.7]{perf-results/poisson}
%%\caption{Poisson results.}\label{fig:poisson}
%\end{figure}
%
%In Figure~\ref{fig:poisson} we report speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the original code for the Poisson form. We note that, as a general trend, \texttt{ffc} exhibits drops in performance as $nf$ increases, notably when $nf=3$, for any values of $q$ and $p$. This is a consequence of the inherent complexity of the generated code. The way \texttt{ffc} performs loop-invariant code motion leads to the pre-computation of integration-dependent terms at the level of the integration loop, which are characterized by higher arithmetic intensity and redundant computation as $nf$ increases. Moreover, the absence of vectorization is another limiting factor.
%
%The \texttt{auto} variant generally shows the best performance. Significant improvements over \texttt{fix} are also achieved, notably as $q$, $p$ and $nf$ increase. As clarified in the next section, this is always due to a more aggressive expression rewriting in combination with the zero-avoidance technique.

>>>>

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache. The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost, -ip}.

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix and the bilinear forms arising in a Helmholtz equation, in an elastic model, and in a hyperelastic model. The complete specification of these forms is made publicly available (\cite{ufl-code}).

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item the FEniCS Form Compiler's built-in optimization systems, namely
\begin{itemize}
\item optimized quadrature mode, \texttt{ffc-quad}
\item tensor mode, \texttt{ffc-tens}
\item automatic mode, \texttt{ffc-auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
\end{itemize}
\item a novel back-end for the FEniCS Form Compiler, UFLACS \texttt{uflacs}, whose primary goals are improved code generation time and runtime
\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{coffee-O1}, as discussed in~\citeN{Luporini}
\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{coffee-O2}, which is the focus of this paper
\end{itemize}

The values that we report are the average of three runs with "warm cache" (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation. For a fair comparison, small patches (publicly available) were written to be able to use the aforementioned transformation systems in Firedrake; that is, all simulations are run through Firedrake. UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS (\cite{Fenics}), which employs a data storage layout different than Firedrake. Our small patches fix this problem by producing code with a "correct" data storage layout.

Following the methodology (and the notation) used in ~\cite{quadrature1} and~\cite{Francis}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis are instead
\begin{itemize}
\item the space of test, trial, and coefficient functions, Lagrange
\item the mesh, tetrahedral with a total of 4374 elements
\end{itemize}

Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will deeply be commented in Section~\ref{sec:perf-results-forms}, can be interpreted as ``plots (grids) of plots''. Each grid (each figure) has two logical "outer" axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. Looking at the diagonal of the grid one can obtain the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid (figure) can therefore be read in many different ways, which allows us to make structured considerations on the performance achieved. In each plot there are three groups of bars, each group referring to a particular version of the code (\texttt{ffc-quad, ffc-tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

\paragraph{Mass}
%\begin{figure}[t]  
%...
%%\includegraphics[scale=0.7]{perf-results/mass}
%%\caption{Mass results.}\label{fig:mass}
%\end{figure}
%

@@
